Correcting for bias in high-throughput sequencing data

### 2011-09-24 seqbias-paper.tex
significance tests, minor tweaks

\begin{table}
\processtable{Data sets on which the methods are evaluated.\label{tab:datasets}}
{
\begin{tabular}{lllr}\toprule
Experiment & Species & Platform & Read Length \\\midrule
\citet{Wetterbom2010} & Chimpanzee & ABI & 33 \\
Katze (unpublished) & Macaque & ABI & 50 \\
\citet{Bullard2010} & Human & Illumina & 35 \\
\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
\end{tabular}
}{}
\end{table}

@@

\begin{figure}
\centerline{\includegraphics[width=0.30\textwidth]{pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we compute a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive, and increases as uniformity is
increased, and negative when uniformity is decreased. Boxes are plotted to mark
the 25\%, 50\%, and 75\% quantiles, with whiskers extending to 1.5 times the
inter-quartile range (i.e., the span between the 25\% and 75\% quantiles), and
dots marking more extreme values.
}

@@

\begin{table}
\processtable{
    The median $R^2$ goodness of fit statistic across test exons.
\label{tab:pois}}{
    \begin{tabular}{lcccc}\toprule
          & BN    & MART  & GLM   & 7mer \\\midrule
Wetterbom & 0.174 & 0.016 & 0.066 & -0.079 \\
Katze     & 0.280 & 0.243 & 0.158 & 0.033 \\
Bullard   & 0.267 & 0.163 & 0.224 & 0.157 \\
Mortazavi & 0.240 & 0.210 & 0.197 & 0.091 \\
Trapnell  & 0.289 & 0.289 & 0.248 & 0.138
    \end{tabular}
}
{
    The $R^2$ statistic measures increased uniformity in read coverage, after
    correcting for bias.  Here the median $R^2$ across the test exons is listed
    for each method and sample. A higher $R^2$ indicates a better fit.
}
\end{table}

@@

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{pcrcor.eps}}
\caption{
    Pearson correlation is computed betweer TaqMan values and (bias-corrected)
    read counts, on a logarithmic scale.
    %%TODO
    Error bars show a 95\% confidence interval, estimated using the Fisher
    transformation, as computed by the R function \texttt{cor.test}.
}
\label{fig:pcrcor}
\end{figure}

%\begin{table}
%\processtable{
    %Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
    %values.\label{tab:pcr}}{
%\begin{tabular}{ll} \toprule
%Method & Correlation \\ \midrule
%Unadjusted &  0.6665 \\
%7mer &  0.6677 \\
%GLM & 0.6854 \\
%MART & 0.6921 \\
%BN & 0.7016
%\end{tabular}
%}
%{
%}
%\end{table}

>>>>

\begin{table}
\processtable{Data sets on which the methods are evaluated.\label{tab:datasets}}
{
\begin{tabular}{lllcc}\toprule
 & & & Read & Num. of \\
Experiment & Species & Platform & length & reads \\\midrule
\citet{Wetterbom2010} & Chimpanzee & ABI & 33 & $3.9 \times 10^7$ \\
Katze (unpublished) & Macaque & ABI & 50  & $4.8 \times 10^7$ \\
\citet{Bullard2010} & Human & Illumina & 35 & $1.2 \times 10^7$ \\
\citet{Mortazavi2008} & Mouse & Illumina & 33 & $3.9 \times 10^6$ \\
\citet{Trapnell2010} & Mouse & Illumina & 75 & $6.6 \times 10^7$ \\\botrule
\end{tabular}
}{}
\end{table}

@@

\begin{figure}
\centerline{\includegraphics[width=0.30\textwidth]{pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we compute a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive, and increases as uniformity is
increased, and negative when uniformity is decreased.  Marked with stars are
methods over which our approach showed a statistically significant improvement
when applied to the same data, according to a one-sided Wilcoxon signed-rank
test. In each of those marked, we observed p-values less than $10^{-23}$.  Boxes
are plotted to mark the 25\%, 50\%, and 75\% quantiles, with whiskers extending
to 1.5 times the inter-quartile range (i.e., the span between the 25\% and 75\%
quantiles), and dots marking more extreme values.
}

@@

\begin{table}
\processtable{
    The median $R^2$ goodness of fit statistic across test exons.
\label{tab:pois}}{
    \begin{tabular}{lcccc}\toprule
          & BN    & MART  & GLM   & 7mer \\\midrule
Wetterbom & \textbf{0.174} & 0.016 & 0.066 & -0.079 \\
Katze     & \textbf{0.280} & 0.243 & 0.158 & 0.033 \\
Bullard   & \textbf{0.267} & 0.163 & 0.224 & 0.157 \\
Mortazavi & \textbf{0.240} & 0.210 & 0.197 & 0.091 \\
Trapnell  & \textbf{0.289} & \textbf{0.289} & 0.248 & 0.138
    \end{tabular}
}
{
    The $R^2$ statistic measures increased uniformity in read coverage, after
    correcting for bias.  Here the median $R^2$ across the test exons is listed
    for each method and sample. A higher $R^2$ indicates a better fit.
}
\end{table}

@@

%\begin{figure}
%\centerline{\includegraphics[width=0.4\textwidth]{pcrcor.eps}}
%\caption{
    %Pearson correlation is computed betweer TaqMan values and (bias-corrected)
    %read counts, on a logarithmic scale.
    %%%TODO
    %Error bars show a 95\% confidence interval, estimated using the Fisher
    %transformation, as computed by the R function \texttt{cor.test}.
%}
%\label{fig:pcrcor}
%\end{figure}

\begin{table}

\processtable{
    Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
    values.\label{tab:pcr}}{
    \centerline{
\begin{tabular*}{0.4\textwidth}{lrl}
\toprule
Method & Correlation & \\ \midrule
Unadjusted &  0.6665 & $\ast\ast$ \\
7mer &  0.6677 & $\ast\ast$ \\
GLM & 0.6854 & $\ast\ast$ \\
MART & 0.6921 & $\ast$ \\
BN & \textbf{0.7016}
\end{tabular*}}

}
{
We estimated the statistical significance of the improvement in correlation
using the BN method over the other methods using a simple boostrap procedure. A
bootstrap sample is formed by sampling, with replacement, 592 genes from the
original set of the same size. The correlation is then computed for this set,
using the adjusted count from each method. We repeated this procedure one
million times, and counted the number of times each of the competing method achieves a
higher correlation than the BN method.  Those marked with a sigle asterisk
achieved a higher correlation fewer than 1000 times, resulting in a p-value of
less than $10^{-3}$ Those marked with two asterisks achieved a higher
correlation in none of the bootstrap samples, indicating a p-value less than
$10^{-6}$.
}
\end{table}
